\large{\bf Large effects and the infinitesimal model }

### 2023-07-04 writeup.tex
edits through intro

\title{\large{\bf
    Normality and large effect loci in quantitative genetics
}}

@@

\begin{abstract}
\end{abstract}

@@

Much of the work to give a quantitative understanding to the observation that
offspring of sexual organisms resemble their parent,
has focused on \emph{quantitative genetics},
which is interestingly intermediate between a mechanistic model of genetics
and a purely phenomenological model.
The action of natural selection on genetic variants in the absence of genetic variation
is relatively well-understood (cite ewens),
but tractable mathematical models of evolution for recombining species
are elusive,
at least with some results of when the model might be a good description of reality.
For instance, a great deal of conceptual understanding and even quantitative predictions
have been made assuming that the distribution of trait values in a population is Gaussian
(cite Lande etcetera)
despite the absence of a plausible model that would make this true (cite turelli and barton).
One of the oldest and most concrete models of heritable trait evolution is the \emph{infinitesimal model}
which assumes a simple, explicit form for how an offspring's trait is inherited from their parents':
it is Normally distributed, with mean equal to the average of the parents' values
and a variance independent of their values.
Despite the plausibility of this model,
(thanks to the Central Limit Theorem),
it was not until \citet{barton2017infinitesimal} that conditions under which this would be a good approximation
were made explicit,
through a formal convergence proof.

As suggested by its name, under the infinitesimal model the effects of any given genetic variant on the trait
is small (indeed, vanishing),
and so selection-driven trait shifts are mediated by small changes in frequencies of many alleles of small effect.
However, empirical understanding of the genetic basis of trait variation in real-world species
often finds a significant portion of heritable variation explained by variants at only a few loci.
There are many possible explanations for this observation,
including selection/publication bias,
filters for larger-effect variants due to spatially varying selection (cite),
ond/or grouping together of concordant alleles by linkage or inversions (cite).
However, it seems clear that in at least those cases where we understand the underlying molecular mechanisms
that mutations of large effect can be important in practice
(cite examples: insecticide resistance, coloration, etc),
and the consensus in the field is that traits exist on a spectrum between monogenic and polygenic,
and that many - if not most - are firmly in the intermediate, ``oligogenic'' region (cite),
where heritable trait variation is due both to segregating large- and small-effect variants.
Indeed, some sensible verbal models of the biological mechanisms by which a genetic change
comes to have an effect on an organism's trait
describe a hierarchy or range of ``proximities'' to the trait,
by which changes to more ``nearby'' genes have the possibility of causing larger changes in the trait
(cite omnigenic and others).
This suggests a mixture model for the effect size distribution:
perhaps the distribution of effect sizes within each gene is Normal,
but with a standard deviation that varies with ``proximity'' to the trait;
such mixtures can easily have large proportions of extreme values.

Concretely, the infinitesimal model postulates that
if the parents' genetic contributions to the trait (i.e., their \emph{breeding values})
are $z^{[1]}$ and $z^{[2]}$,
and the offspring's genetic contribution to the trait is $z$,
and $R = z - (z^{[1]} + z^{[2]})/2$ is the deviation from the midparent
(which we call the \emph{segregation noise}), then
\begin{enumerate}
    \item $R$ is independent of $z^{[1]}$ and $z^{[2]}$, and
    \item $R$ has a Gaussian distribution.
\end{enumerate}
(This is not a full definition,
since the model makes broader statements involving conditioning on the full pedigree;
see \citet{barton2017infinitesimal}.)
We set out here to see whether (2) can be generalized,
and find that (1) is apparently inexitrably connected.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality
of adding up a great many very small and independent things.
However, as is well known to probabilists,
there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \} \propto t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, the $\alpha$-stable distribution (and becomes exact, as $n \to \infty$).
In such ensembles, sinle large entries are important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.

This suggests exploring whether stable distributions might reasonably stand in for the Gaussian
in the infinitesimal model.
In this paper, we make preliminary explorations in that direction:
Does the universality shown by \citet{barton2017infinitesimal}
carry over when the distribution of effect sizes falls in the domain of attraction of a stable law?
What can be said about equilibrium trait distributions in this case?
And, is there any biological evidence that this theory might be relevant?

Stable distributions are by no means new to population biology.
For instance, it is a long-standing question how often the motions of organisms
are better modeled by essentially Brownian random walks or by L\'evy processes (cite),
and a growing literature studies the effects of such ``fat-tailed'' distributions
on the shapes and dynamics of traveling waves such as biological invasions (cite).
\comment{Other examples?}
More directly analogous to the problem we study here, \citet{landis2012phylogenetic}
found that an $\alpha$-stable process (with $\alpha \approx 1.5$) fit the evolution of log endocranial volume
on the primate phylogeny
better than either a Brownian model.
\comment{more}

\comment{TODO expand, put somewhere:}
There are several justifications for why a ``stable infinitesimal model'' might be useful:
(1) as a phenomenological model;
(2) if it matches empirical observations; or
(3) if there is a central limit theorem suggesting it would be a good approximation in a wide range of situations.

\subsection*{Orphaned text}
The ways in which genetic variants combine to determine an organisms' trait can be complex,
and epistatic interactions between alleles and between loci can be important in practice.
However, the question is sufficiently complex in the purely additive case
that a trait can be decomposed into a sum across loci of the contributions of the alleles from each locus.
Under this additive assumption it makes sense to discuss the ``effect size distribution'',
i.e., the distribution of how much alleles affect the trait value.

An offspring's trait differs from their parents for many reasons,
often most importantly because of differences in the environment in which they live.
The focus of quantitative genetics is on the contribution of heritable material to this variation:
the offspring's genetic material

\comment{Will need this notation below:}
We focus on the \emph{infinitesimal model},
as described by \citet{barton2017infinitesimal},
which supposes that the trait value of an offspring
is equal to the average of their parents' trait values,
plus independent Gaussian noise with fixed variance,
\begin{equation} \label{eqn:offspring_model}
    X_o = \frac{X_m + X_f}{2} + \epsilon,
\end{equation}
where $\epsilon$, the ``segregation noise'',
is Gaussian with mean zero and variance $V_0$.
\comment{should call this something besides $\epsilon$}

@@

We applied this procedure to results from models fit on both sexes,
after removed phenotypes related to employment, drug treatments, and diet.
This left us with 1,108 phenotypes for which which could estimate the tail exponent.
For the main results,
we then restricted to phenotypes having non-missing numbers of ``cases'' and ``controls''
and a sample size of at least 300,000 (i.e., not more than 61,194 of 361,194 missing phenotypes).
We also removed
the 65 remaining phenotypes that are common (more than 10,000 cases),
as these often showed very different patterns
(e.g., effect size distributions).
This left us with 695 phenotypes;
plots including the 413 ``filtered'' phenotypes are provided in
Supplementary Figure~\ref{sfig:unfiltered_hist}.

@@

\begin{lem}\label{lem:parental_contribs}
    Suppose that $\{\eta_\ell^{[i]}\}$ are independent for all $\ell$ and $i$,
    that the distribution of $\eta_\ell^{[i]}$ only depends on $\ell$,
    and that $\{X_\ell\}$ are independent of $\{\eta_\ell^{[i]}\}$.
    If the segregation noise is independent of the midparent value
    (i.e., $R_M$ is independent of $(Z^{[1]} + Z^{[2]})/2$)
    then $R_M$ is Gaussian.
\end{lem}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    Segregation noise with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
    from the same simulations shown in Figure~\ref{fig:trait_distrns},
    for all individuals born during the last 1,000 time units of the simulation.
    \textbf{(A,D)} Segregation noise (offspring trait minus midparent) plotted against midparent.
    \textbf{(B,E)} Histograms of segregation noise.
    \textbf{(B,E)} Enrichment in quantiles of segregation noise conditional on midparent value.
    To compute the latter,
    let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of segregation noise
    into 50 bins with roughly equal numbers in each (i.e., the quantiles),
    let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
    is in $[a, b)$ whose segregation noise is in $[q_{k-1}, q_k)$,
    and let $p_k$ be the proportion of all offspring with segreation noise in $[q_{k-1}, q_k)$
    (so, $p_k \approx 0.02$).
    Then, each line shows $p_k(a,b)/p_k$
    plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
    for $(a,b)$ chosen to span the 5\% of midparents centered on:
    (red) the 10\% quantile,
    (black) the median, and
    (green) the 90\% quantile of midparent value.
    \caption{
        \label{fig:seg_noise}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.1.
        \label{fig:seg_noise_small}
    }
\end{figure}

@@

@@

\appendix

@@

>>>>

\title{\large{\bf
    Large effects and the infinitesimal model
}}

@@

\begin{abstract}
    The infinitesimal model of quantitative genetics
    relies on the Central Limit Theorem to stipulate that
    under additive models of quantitative traits
    determined by many loci having similar effect size,
    the difference between an offspring's genetic trait component
    and the average of their two parents' genetic trait components
    is Normally distributed and independent of the parents' values.
    Here, we investigate how the assumption of similar effect sizes
    affects the model:
    in particular, if the tail of the effect size distribution is polynomial
    with exponent $\alpha < 2$, then sums of effects
    should be well-approximated by a ``stable distribution'',
    and find tail exponents between 1 and 2 in 
    effect sizes estimated by genome-wide association studies
    of many human disease-related traits.
    We show that the independence of offspring trait deviations from parental averages
    in many cases implies a Gaussian distribution,
    suggesting that non-Gaussian models of trait evolution must expicitly
    track the underlying genetics, at least for loci of large effect.
    We also characterize possible limiting trait distributions
    of the infinitesimal model with infinitely divisible noise distributions,
    and compare our results to simulations.
\end{abstract}

@@

Much of the work to 
understand the evolution of quantitative traits in sexual species
has focused on models in quantitative genetics
that describe evolution of trait values without explicitly representing underlying genetic variation,
but in a manner deriving from our understanding of how genetic variation works in practice.
Such models are therefore
interestingly intermediate between mechanistic models of genetics and purely phenomenological.
% The action of natural selection on genetic variants in the absence of genetic variation is relatively well-understood (cite ewens),
% but tractable mathematical models of evolution for recombining species are elusive,
% at least with some results of when the model might be a good description of reality.
% For instance, a great deal of conceptual understanding and even quantitative predictions
% have been made assuming that the distribution of trait values in a population is Gaussian (cite Lande etcetera)
% despite the absence of a plausible model that would make this true (cite turelli and barton).
One concrete model of heritable trait evolution is the \emph{infinitesimal model},
which assumes a simple, explicit form for how the genetic component of offsprings' traits are determined by their parents:
they are Normally distributed, with the mean for each offspring equal to the average of the parents' values,
and a covariance independent of the parental trait values (but that depends on how they are related).
\citet{barton2017infinitesimal} give a history of the model
and a formal proof of conditions under which this would be a good approximation to reality,
which they extended to include dominance in \citet{barton2022infinitesimal}.

\comment{TODO: citations below}

As suggested by its name,
under the infinitesimal model the effects of any given genetic variant on the trait
is small (indeed, vanishing),
and selection-driven trait shifts are mediated by small changes in frequencies of many alleles of small effect.
However, empirical understanding of the genetic basis of trait variation in real-world species
often finds a significant portion of heritable variation explained by variants at only a few loci.
There are many possible explanations for this observation,
including selection or publication bias,
filters for larger-effect variants due to spatially varying selection (cite),
and/or grouping together of concordant alleles by linkage or inversions (cite).
However, it seems clear that in at least those cases where we understand the underlying molecular mechanisms,
mutations of large effect can be important in practice
(cite examples: insecticide resistance, coloration, etcetera).
The consensus in the field seems to be that traits exist on a spectrum between monogenic and polygenic,
and that many - if not most - are firmly in the intermediate, ``oligogenic'' region (cite),
where heritable trait variation is due both to segregating large- and small-effect variants.
Indeed, some sensible verbal models of the biological mechanisms by which a genetic change
comes to have an effect on an organism's trait
describe a hierarchy or range of ``proximities'' to the trait,
by which changes to more ``nearby'' genes have the possibility of causing larger changes in the trait
(cite omnigenic and others).
This suggests a mixture model for the effect size distribution:
perhaps the distribution of effect sizes within each gene is Normal,
but with a standard deviation that varies with ``proximity'' to the trait;
such mixtures can easily have large proportions of extreme values.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality
of adding up a great many very small and independent things.
However, there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \}$ is proportional to $t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, an $\alpha$-stable distribution
(and the approximation becomes exact, as $n \to \infty$).
In such ensembles, single large entries can be important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.

This suggests exploring whether stable distributions might reasonably stand in for the Gaussian
in the infinitesimal model.
In this paper, we make preliminary explorations in that direction:
Do the results of \citet{barton2017infinitesimal}
carry over when the distribution of effect sizes falls in the domain of attraction of a stable law?
The immediate answer to this will be ``no'' --
in fact, we will see that independence of offsprings' deviations and midparent values \emph{implies} a Gaussian distribution in some sense.
This independence is a key reason why the infinitesimal model is tractable,
and so a conclusion is that additive models with non-Gaussian effects may not be well-approximated
by ``trait-only'' models that do not explicitly represent underlying genetic variation.
Having thus shown that ``infinitesimal models with non-Gaussian noise'' is on weak footing as a model of reality,
we then go on to analyze exactly these models.
(As motivation,
we have not shown that such models are \emph{not} good approximations to reality,
and it is interesting and perhaps informative to understand how the noise distribution affects results.)

Stable distributions are by no means new to population biology.
For instance, it is a long-standing question how often the motions of organisms
are better modeled by essentially Brownian random walks or by L\'evy processes (cite),
and a growing literature studies the effects of such power law distributions
on the shapes and dynamics of traveling waves such as biological invasions (cite).
\comment{Other examples?}
More directly analogous to the problem we study here, \citet{landis2012phylogenetic}
found that an $\alpha$-stable process (with $\alpha \approx 1.5$) fit the evolution of log endocranial volume
on the primate phylogeny
better than either a Brownian model.
It is also important to note that
the infinitesimal model implies a Gaussian distribution for the difference between an offspring's trait and their parents' average,
\emph{not} for the entire population.
The question of when the trait distribution in an evolving population
might be modeled as a Gaussian has been discussed extensively, e.g., by \comment{cite}.

The ways in which genetic variants affect an organisms' trait can be complex,
and interactions between alleles, between loci, and with the environment are often important in practice.
In fact, many traits claimed to be genetically determined
in fact only appear heritable through non-genetic means,
e.g., systematic inequalities in resource distribution partitioned by class or race. \comment{e.g., cite}
In this paper, we follow the literature in working with the relatively simple purely additive case
of a trait that can be decomposed into a sum across loci of the contributions of the alleles from each locus,
even omitting an independent effect of ``the environment''.
However, we emphasize that such models should \emph{not} always be the default for understanding biological variation,
especially for human traits.

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Notation}

We generally follow the notation of \citet{barton2017infinitesimal},
although in a simpler situation,
as we do not consider full pedigrees
but rather only the haploid offspring of a single set of haploid parents
(as produced, for instance, by a round of meiosis in a diploid stage),
and ignore non-genetic contributions to the trait;
so, below we say ``trait'' to mean ``genetic contribution to the trait''.

Suppose that the focal individual's trait is
$$
    Z_M = \sum_{\ell=1}^M \eta_\ell / M^{1/\alpha},
$$
where $\eta_\ell$ is the contribution of the allele carried at the $\ell^\text{th}$ locus.
The effects of the alleles carried by the two parents
are $\eta_\ell^{[1]}$ and $\eta_\ell^{[2]}$, respectively,
and their traits are
$Z_M^{[1]} = \sum_\ell \eta_\ell^{[1]} / M^{1/\alpha}$ and
$Z_M^{[2]} = \sum_\ell \eta_\ell^{[2]} / M^{1/\alpha}$.
We assume the loci are unlinked and inheritance is fair, so
letting $X_\ell$ be a family of Bernoulli(1/2) random variables that are independent of everything else,
$$
    \eta_\ell = X_\ell \eta_\ell^{[1]} + (1 - X_\ell) \eta_\ell^{[2]} .
$$
The \emph{transmission noise}, $R_M$, is the offspring's trait minus the midparent value, $\bar Z_M$:
\begin{align*}
    \bar Z_M &= \frac{Z_M^{[1]} + Z_M^{[2]}}{2} \\
    R_M &= Z_M - \bar Z_M \\
        &= \sum_{\ell=1}^M
            \frac{
                X_\ell (\eta_\ell^{[1]} - \eta_\ell^{[2]})
                + (1 - X_\ell) (\eta_\ell^{[2]} - \eta_\ell^{[1]})
            }{ 2 } .
\end{align*}

@@

We applied this procedure to results from models fit on both sexes,
after removed phenotypes related to employment, drug treatments, and diet.
This left us with 1,108 phenotypes for which which could estimate the tail exponent.
For the main results,
we then restricted to phenotypes having non-missing numbers of ``cases'' and ``controls''
and a sample size of at least 300,000 (i.e., not more than 61,194 of 361,194 missing phenotypes).
We also removed
the 65 remaining phenotypes that are common (more than 10,000 cases),
as these often showed very different patterns
(e.g., effect size distributions).
This left us with 695 phenotypes;
plots including the 413 ``filtered'' phenotypes are provided in
Supplementary Figure~\ref{fig:unfiltered_hist}.

@@

\begin{lem}\label{lem:parental_contribs}
    Suppose that $\{\eta_\ell^{[i]}\}$ are independent for all $\ell$ and $i$,
    that the distribution of $\eta_\ell^{[i]}$ only depends on $\ell$,
    and that $\{X_\ell\}$ are independent of $\{\eta_\ell^{[i]}\}$.
    If the segregation noise is independent of the midparent value
    (i.e., $R_M$ is independent of $\bar Z_M$)
    then $R_M$ is Gaussian.
\end{lem}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    \caption{
        Segregation noise with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
        from the same simulations shown in Figure~\ref{fig:trait_distrns},
        for all individuals born during the last 1,000 time units of the simulation.
        \textbf{(A,D)} Segregation noise (offspring trait minus midparent) plotted against midparent.
        \textbf{(B,E)} Histograms of segregation noise.
        \textbf{(B,E)} Enrichment in quantiles of segregation noise conditional on midparent value.
        To compute the latter,
        let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of segregation noise
        into 50 bins with roughly equal numbers in each (i.e., the quantiles),
        let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
        is in $[a, b)$ whose segregation noise is in $[q_{k-1}, q_k)$,
        and let $p_k$ be the proportion of all offspring with segreation noise in $[q_{k-1}, q_k)$
        (so, $p_k \approx 0.02$).
        Then, each line shows $p_k(a,b)/p_k$
        plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
        for $(a,b)$ chosen to span the 5\% of midparents centered on:
        (red) the 10\% quantile,
        (black) the median, and
        (green) the 90\% quantile of midparent value.
        \label{fig:seg_noise}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.1.
        The same quantities for the simulations of Figure~\ref{fig:sel_seg_noise}
        with stabilizing selection are shown in Figure~\ref{fig:sel_seg_noise_small}.
        \label{fig:seg_noise_small}
    }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Not assuming all effects are the same order of magnitude seems very bioloically realistic,
and so it'd be nice to figure out how evolution works and how to do quant gen in such cases.
Such models can always be simulated,
but the way forwards with theoretical understanding seems murky.
It would be interesting to see how changing the distribution of segregation noise
in the infinitesimal model
affects properties like the rate of adaptive evolution,
levels of genetic load,
and the rate of fixation.
However,
we have seen that this assumption imples a sort of ``blending inheritance''
for the large effect alleles,
and so such work much be accompanied by simulations
to see if the predictions are borne out under concrete models of genetic inheritance.
Another open question is whether there is some set of assumptions
(perhaps involving linkage, pleiotropy, and/or environmental interactions with genetic effects)
which leads to these non-Gaussian infinitesimal models.
Finally, it may well be that there is a clever way to set up a trait-only model of evolution
that is both insensitive to the underlying genetics
and does not rely on the Gaussian central limit theorem.

Above we've mostly used the Cauchy (e.g., in simulations)
not because it seems most realistic
(indeed data suggests $\alpha=3/2$ might be better)
but for convenience and to make effects more obvious.

@@

\appendix
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.1.
        \label{fig:sel_seg_noise_small}
    }
\end{figure}
